---
title: "Pertussis Resurgence"
author: "Vina Nguyen"
date: "2023-06-12"
output: html_document
---

## Mini-Project
Investigating Pertussis Resurgence

Pertussis (more commonly known as whooping cough) is a highly contagious respiratory disease caused by the bacterium Bordetella pertussis. People of all ages can be infected leading to violent coughing fits followed by a high-pitched intake of breath that sounds like “whoop”. Infants and toddlers have the highest risk for severe complications and death. Recent estimates from the WHO suggest that ~16 million cases and 200,000 infant deaths are due to pertussis annually 1.

Bordetella pertussis attacks cells lining the airways. The rope-like structures shown are cilia, that typically sweep away inhaled dirt and foreign objects. In a pertussis infection, the bacteria use adhesive proteins to stick whilst releasing toxins that damage cells, trigger inflammation and increase mucus production leading to uncontrollable violent coughing.

### 1. Investigating pertussis cases by year
The United States Centers for Disease Control and Prevention (CDC) has been compiling reported pertussis case numbers since 1922 in their National Notifiable Diseases Surveillance System (NNDSS). We can view this data on the CDC website here: https://www.cdc.gov/pertussis/surv-reporting/cases-by-year.html

> Q1. With the help of the R “addin” package datapasta assign the CDC pertussis case number data to a data frame called cdc and use ggplot to make a plot of cases numbers over time.

```{r}
library(datapasta)
```

```{r}

```











































































## what

















































