---
title: "class17 Vaccination Ratee Mini-Project EC"
author: "Vina Nguyen"
date: "2023-06-09"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

### COVID-19 Vaccination Rates
The goal of this hands-on mini-project is to examine and compare the Covid-19 vaccination rates around San Diego.

We will start by downloading the most recently dated “Statewide COVID-19 Vaccines Administered by ZIP Code” CSV file from: https://data.ca.gov/dataset/covid-19-vaccine-progress-dashboard-data-by-zip-code

```{r}
# import vaccination data
vax <- read.csv("vax.csv")
head(vax)
```

```{r}
vax$as_of_date
```


> Q1. What column details the total number of people fully vaccinated? 

persons_fully_vaccinated

> Q2. What column details the Zip code tabulation area? 

zip_code_tabulation_area

> Q3. What is the earliest date in this dataset? 

2021-01-05

> Q4. What is the latest date in this dataset?

2023-06-06

```{r}
# skim fuction to get quick overview of dataset
skimr::skim(vax)
```

## Getting Started

> Q5. How many numeric columns are in this dataset? 

14

> Q6. Note that there are “missing values” in the dataset. How many NA values there in the persons_fully_vaccinated column? 

```{r}
# is.na function indicates which elements are missing
sum( is.na(vax$persons_fully_vaccinated) )
```

> Q7. What percent of persons_fully_vaccinated values are missing (to 2 significant figures)? 

```{r}
(sum( is.na(vax$persons_fully_vaccinated) )) / (nrow(vax))

```
0.08 8%

> Q8. [Optional]: Why might this data be missing?

Data was not taken in certain regions or was not updated. 


One of the “character” columns of the data is as_of_date, which contains dates in the Year-Month-Day format.

Dates and times can be annoying to work with at the best of times. However, in R we have the excellent lubridate package, which can make life allot easier. Here is a quick example to get you started:

```{r}
library(lubridate)
```

What is today's date?
```{r}
today()
```

The as_of_date column of our data is currently not that usable. For example we can’t easily do math with it like answering the simple question how many days have passed since data was first recorded.

```{r}
# This will give an Error!
today() - vax$as_of_date[1]
```

However if we convert our date data into a lubridate format things like this will be much easier as well as plotting time series data later on.

```{r}
# Specify that we are using the year-month-day format
vax$as_of_date <- ymd(vax$as_of_date)
```

Now we can do math with dates. For example: How many days have passed since the first vaccination reported in this dataset?

```{r}
today() - vax$as_of_date[1]
```

Using the last and the first date value we can now determine how many days the dataset span?

```{r}
vax$as_of_date[nrow(vax)] - vax$as_of_date[1]
```

> Q9. How many days have passed since the last update of the dataset? 

```{r}
today() - vax$as_of_date[224028]
```

> Q10. How many unique dates are in the dataset (i.e. how many different dates are detailed)? 



## Working with ZIP codes

```{r}
library(zipcodeR)
```

```{r}
geocode_zip('92037')
```

Calculate the distance between the centroids of any two ZIP codes in miles, e.g.

```{r}
zip_distance('92037','92109')
```

More usefully, we can pull census data about ZIP code areas (including median household income etc.). For example:

```{r}
reverse_zipcode(c('92037', "92109") )
```

## Focus on the San Diego area

Let’s now focus in on the San Diego County area by restricting ourselves first to vax$county == "San Diego" entries. We have two main choices on how to do this. The first using base R the second using the dplyr package:

```{r}
# Subset to San Diego county only areas
sd <- vax[vax$county, "San Diego" ]
```

Using dplyr the code would look like this:

```{r}
library(dplyr)

sd <- filter(vax, county == "San Diego")

nrow(sd)
```

Using dplyr is often more convenient when we are subsetting across multiple criteria - for example all San Diego county areas with a population of over 10,000.

```{r}
sd.10 <- filter(vax, county == "San Diego" &
                age5_plus_population > 10000)
```

> Q11. How many distinct zip codes are listed for San Diego County? 



>Q12. What San Diego County Zip code area has the largest 12 + Population in this dataset? 











