---
title: "week08"
author: "Vina Nguyen"
date: "2023-05-21"
output: pdf_document
---

## Section 4: Population Scale Analysis

One sample is obviously not enough to know what is happening in a population. You are 
interested in assessing genetic differences on a population scale. So, you processed about ~230 samples and did the normalization on a genome level. Now, you want to find whether there is any association of the 4 asthma-associated SNPs (rs8067378…) on ORMDL3 expression.

This is the final file you got ( https://bioboot.github.io/bggn213_W19/classmaterial/rs8067378_ENSG00000172057.6.txt ). The first column is sample name, the second 
column is genotype and the third column are the expression values.

> Q13: Read this file into R and determine the sample size for each genotype and their corresponding median expression levels for each of these genotypes.

How many samples do we have?

```{r}
expr <- read.table("rs8067378_ENSG00000172057.6.txt")
head(expr)
```

```{r}
nrow(expr)
```

How many of each type?

```{r}
table(expr$geno)
```

```{r}
library(ggplot2)
```

> Q14: Generate a boxplot with a box per genotype, what could you infer from the relative expression value between A/A and G/G displayed in this plot? Does the SNP effect the expression of ORMDL3?

Lets make a boxplot.

```{r}
exp_geno <- ggplot(expr) +
  aes(geno, exp, fill=geno) +
  geom_boxplot(notch=TRUE)
exp_geno
```

Expression with AA is highest. Expression with GG is lowest. Having a GG in this location is associated with having a reduced expression of this gene. The SNP effects the expression of ORMDL3 since it is in the chromosome 17 location with G alleles.


























