---
title: "Machine Learning 1"
author: "Vina Nguyen"
date: "2023-05-01"
output:
  pdf_document: default
  html_document: default
---

## kmeans()

Demo of using kmeans() function in base R. First make up some data with a known structure. 

```{r}
tmp <- c(rnorm(30, -3), rnorm(30, 3))
x <- cbind(x=tmp, y=rev(tmp))
x
```

```{r}
plot(x)
```

Now we have some made up data in 'x' let's see hpw kmeans works with this data.

```{r}
k <- kmeans(x, centers=2, nstart =20)
k
```
> Q. How many points are in each cluster?

```{r}
k$size
```

> Q. How do we get to the cluster membership/assignment?

```{r}
k$cluster
```

> Q. What about cluster centers?

```{r}
k$centers
```

Now we got to the main results. Let's use them to plot out data with the kmeans result.

```{r}
plot(x, col=k$cluster)
points(k$centers, col="blue", pch=15)
```

## hclust()

We will cluster the same data 'x' with the 'hclust()'. In this case, 'hclust()' requiresd a distaznce matrix as input.

```{r}
hc <- hclust( dist(x))
hc
```

Let's plot out hclust results.

```{r}
plot(hc)
```

To get out cluster membership vector, we need to "cut' the tre with the 'cutree()'.

```{r}
grps <- cutree(hc, h=8)
grps
```

Now plot out data with the hclust() results.

```{r}
plot(x, col=grps)
```



# Principal Components Analysis (PCA)

## PCA of UK food Data

Read data from website and try a few visualizations.

```{r}
url <- "https://tinyurl.com/UK-foods"
x <- read.csv(url, row.names=1)
x
```

> Q1. How many rows and columns are in your new data frame named x? What R functions could you use to answer this questions?

```{r}
dim(x)
```
> Q2. Which approach to solving the ‘row-names problem’ mentioned above do you prefer and why? Is one approach more robust than another under certain circumstances?

use argument setting row.names=1.

```{r}
cols <- rainbow(nrow(x))
barplot(as.matrix(x), col=cols)
```

> Q3: Changing what optional argument in the above barplot() function results in the following plot?

use beside=FALSE in barplot() code.

```{r}
barplot(as.matrix(x), col=cols, beside=TRUE)
```
 
```{r}
pairs(x, col=cols)
```

> Q5: Generating all pairwise plots may help somewhat. Can you make sense of the following code and resulting figure? What does it mean if a given point lies on the diagonal for a given plot?

Looking at the top left plots, everything on the x-axis represents England while everything on the y-axis represents Wales. If the value of the given point lies on the diagonal, it means that the values are the same for the x and y-axis.

> Q6. What is the main differences between N. Ireland and the other countries of the UK in terms of this data-set?

N. Ireland points are more off (below) the diagonal.

The main base R PCA function is called 'prcomp()' and we will need to give it the transpose of our input data.

```{r}
pca <- prcomp( t(x))
pca
```

There is a nice summary of how well PCA is doing.
```{r}
summary(pca)
```


```{r}
attributes(pca)
```
> Q7. Complete the code below to generate a plot of PC1 vs PC2. The second line adds text labels over the data points. Q8. Customize your plot so that the colors of the country names match the colors in our UK and Ireland map and table at start of this document.


To make our new PCA plot (aka PCA score plot) we access 'pca$x'.

```{r}
country_cols <- c("orange", "red", "blue", "green")
plot(pca$x[,1], pca$x[,2], xlab="PC1", ylab="PC2")
text(pca$x[,1], pca$x[,2], colnames(x), col=country_cols)
```
```{r}
v <- round( pca$sdev^2/sum(pca$sdev^2) * 100 )
v
```
```{r}
## or the second row here...
z <- summary(pca)
z$importance
```

```{r}
barplot(v, xlab="Principal Component", ylab="Percent Variation")
```

```{r}
## Lets focus on PC1 as it accounts for > 90% of variance 
par(mar=c(10, 3, 0.35, 0))
barplot( pca$rotation[,1], las=2 )
```

```{r}
barplot( pca$rotation[,2], las=2 )
```

> Q9: Generate a similar ‘loadings plot’ for PC2. What two food groups feature prominantely and what does PC2 maninly tell us about?

Fresh_potatoes and Other_Veg are the largest positive loading scores.

```{r}
## The inbuilt biplot() can be useful for small datasets 
biplot(pca)
```


## PCA of RNA-Seq Data

Read in data from website.

```{r}
url2 <- "https://tinyurl.com/expression-CSV"
rna.data <- read.csv(url2, row.names=1)
head(rna.data)
```
> Q10: How many genes and samples are in this data set?

```{r}
dim(rna.data)
nrow(rna.data) #genes
ncol(rna.data) #samples
```

```{r}
pca <- prcomp( t(rna.data), scale=TRUE)
summary(pca)
```

Do our PCA plot of this RNA-Seq.

```{r}
# plot of pc1 and pc2
plot(pca$x[,1], pca$x[,2], xlab="PC1", ylab="PC2")
```

Let label the plot!

```{r}
plot(pca$x[,1], pca$x[,2], xlab="PC1", ylab="PC2")
text(pca$x[,1], pca$x[,2], colnames(rna.data))
```

Scree-plot:

```{r}
plot(pca, main="Quick scree plot")
```

```{r}
## Variance captured per PC 
pca.var <- pca$sdev^2

## Percent variance is often more informative to look at 
pca.var.per <- round(pca.var/sum(pca.var)*100, 1)
pca.var.per
```

```{r}
## A vector of colors for wt and ko samples
colvec <- colnames(rna.data)
colvec[grep("wt", colvec)] <- "red"
colvec[grep("ko", colvec)] <- "blue"

plot(pca$x[,1], pca$x[,2], col=colvec, pch=16,
     xlab=paste0("PC1 (", pca.var.per[1], "%)"),
     ylab=paste0("PC2 (", pca.var.per[2], "%)"))

text(pca$x[,1], pca$x[,2], labels = colnames(rna.data), pos=c(rep(4,5), rep(2,5)))
```

Making a ggplot using the ggplot2 package

```{r}
library(ggplot2)

df <- as.data.frame(pca$x)

# Our first basic plot
ggplot(df) + 
  aes(PC1, PC2) + 
  geom_point()
```

```{r}
# Add a 'wt' and 'ko' "condition" column
df$samples <- colnames(rna.data) 
df$condition <- substr(colnames(rna.data),1,2)

p <- ggplot(df) + 
        aes(PC1, PC2, label=samples, col=condition) + 
        geom_label(show.legend = FALSE)
p
```

```{r}
p + labs(title="PCA of RNASeq Data",
       subtitle = "PC1 clealy seperates wild-type from knock-out samples",
       x=paste0("PC1 (", pca.var.per[1], "%)"),
       y=paste0("PC2 (", pca.var.per[2], "%)"),
       caption="Class example data") +
     theme_bw()
```

